---
title: "A Brief Introduction to Vegan"
author: Gavin L. Simpson
date: CSEE 2015 • May 20th 2015
fontsize: 10pt
classoption: "compress, aspectratio=169"
output:
  beamer_presentation:
    theme: m
    keep_tex: true
    highlight: tango
    slide_level: 2
    template: ./slides-template.tex
    fig_width: 6
    fig_height: 3.5
    fig_crop: false
    pandoc_args: [
      "--latex-engine=xelatex"
    ]
---

```{r setup-options, echo = FALSE, results = "hide", message = FALSE}
knitr::opts_chunk$set(comment=NA, fig.align = "center",
                      out.width = "0.7\\linewidth",
                      echo = TRUE, message = TRUE, warning = TRUE,
                      cache = TRUE)
knitr::knit_hooks$set(crop.plot = knitr::hook_pdfcrop)
```

```{r packages, echo = FALSE, results = "hide", message = FALSE}
library("vegan")
data(varespec)
data(varechem)
```

# Basic Ordination

## Unconstrained ordination

What is \alert{unconstrained}

First we look for major variation, then relate it to environmental variation

vs. constrained ordination, where we only want to see what can be explained by environmental variables of interest

How well do we explain the main patterns in the species data vs how large are the patterns we can expain with the measured data

## Examples of unconstrained ordination

* Principal Components Analysis --- PCA
* Correspondance Analysis --- CA
* Nonmetric Multidimensional Scaling --- NMDS

## Before we get started

Housekeeping
```{r housekeeping, eval=FALSE}
setwd("your/working/dir")
```

```{r loading-1, results='hide', message = FALSE}
library("vegan")
data(dune)
data(dune.env)
```
Data from: Jongman, R.H.G, ter Braak, C.J.F & van Tongeren, O.F.R. (1987). Data Analysis in Community and Landscape Ecology. Pudoc, Wageningen.

## Before we get started | species

```{r loading-2}
dim(dune)                               # number of samples, species
head(dune[,1:6])
```

## Before we get started | environment

\scriptsize
```{r loading-3}
head(dune.env, n=3)
summary(dune.env)
```
\normalsize

## Basic ordination

PCA finds linear combinations of the variables that explain the largest amounts of variance in the data

\scriptsize
```{r pca1}
(pca <- rda(dune))
```

## Basic ordination

Vegan has a wrapper function for doing NMDS ordinations using best practices:

* `metaMDS()`

This will do handy things

- standardize your data if necessary
- perform rotation to PCs
- scale coordinates in half change units

## Basic ordination and plotting

```{r NMDS1, results='hide'}
dune.bray.ord <- metaMDS(dune, distance = "bray", k = 2, trymax = 50)
```

## Basic ordination and plotting (using all defaults)

```{r NMDS2, fig.height = 5, crop.plot = TRUE, out.width = "0.5\\linewidth"}
plot(dune.bray.ord)
```

## Basic ordination and plotting (just plots)

```{r NMDS3, fig.height = 5, crop.plot = TRUE, out.width = "0.5\\linewidth"}
plot(dune.bray.ord, display = "sites")
```

## Basic ordination and plotting (just species)

```{r NMDS4, fig.height = 5, crop.plot = TRUE, out.width = "0.5\\linewidth"}
plot(dune.bray.ord, display = "species")
```

## Site names instead of points

```{r NMDS5, fig.height = 5, crop.plot = TRUE, out.width = "0.5\\linewidth"}
plot(dune.bray.ord, display = "sites", type = "text")
```

## Site names instead of points

\scriptsize
```{r NMDS5-2, fig.height = 5, crop.plot = TRUE, out.width = "0.5\\linewidth"}
plot(dune.bray.ord, display = "sites")
set.seed(314) ## make reproducible
ordipointlabel(dune.bray.ord, display = "sites", scaling = 3, add = TRUE)
```
\normalsize

## Site names instead of points

\scriptsize
```{r NMDS5-5, fig.height = 5, crop.plot = TRUE, out.width = "0.5\\linewidth"}
plot(dune.bray.ord, display = "species")
set.seed(314) ## make reproducible
ordipointlabel(dune.bray.ord, display = "species", scaling = 3, add = TRUE)
```
\normalsize

## Site names instead of points

\scriptsize
```{r NMDS5-6, fig.height = 5, crop.plot = TRUE, out.width = "0.5\\linewidth"}
plot(dune.bray.ord)
set.seed(314) ## make reproducible
ordipointlabel(dune.bray.ord, scaling = 3, add = TRUE)
```
\normalsize

## Larger points

```{r NMDS6, fig.height = 5, crop.plot = TRUE, out.width = "0.5\\linewidth"}
plot(dune.bray.ord, display = "sites", cex=2)
```

## Modifying the display of the points with environmental data 

* Color
* Shape
* Size

## Modifying the color of points

\tiny
```{r NMDS7, fig.height = 5, crop.plot = TRUE, out.width = "0.5\\linewidth"}
colors.vec <- c("red", "blue", "orange", "grey")
plot(dune.bray.ord, display = "sites", type = "n")
points(dune.bray.ord, display = "sites", cex=2, pch = 21, 
       col = colors.vec[dune.env$Management], 
       bg = colors.vec[dune.env$Management])
legend("topright", legend = levels(dune.env$Management), bty = "n",
                      col = colors.vec, pch = 21, pt.bg = colors.vec)
```
\normalsize

## Adding other layers

\scriptsize
```{r NMDS11, fig.height = 5, crop.plot = TRUE, out.width = "0.5\\linewidth"}
plot(dune.bray.ord, display = "sites", cex=2) # just site points
```
\normalsize

## Adding other layers

\scriptsize
```{r NMDS12, fig.height = 5, crop.plot = TRUE, out.width = "0.5\\linewidth"}
plot(dune.bray.ord, display = "sites", cex=2)
ordihull(dune.bray.ord,groups = dune.env$Management, label = TRUE) # convex hulls
```
\normalsize

## Adding other layers

\scriptsize
```{r NMDS14, fig.height = 5, crop.plot = TRUE, out.width = "0.5\\linewidth"}
plot(dune.bray.ord, display = "sites", cex=2)
ordihull(dune.bray.ord,groups = dune.env$Management, label = TRUE, col = "blue")
ordispider(dune.bray.ord,groups = dune.env$Management, label = TRUE)
```
\normalsize

## Adding other layers - axes scaling

\scriptsize
```{r NMDS17, fig.height = 5, crop.plot = TRUE, out.width = "0.5\\linewidth"}
plot(dune.bray.ord, type = "n")
```
\normalsize

## Adding other layers - axes scaling

\scriptsize
```{r NMDS18, fig.height = 5, crop.plot = TRUE, out.width = "0.5\\linewidth"}
plot(dune.bray.ord, type = "n")
points(dune.bray.ord, display = "sites", cex = 2)
```
\normalsize

## Adding other layers - axes scaling

\scriptsize
```{r NMDS19, fig.height = 5, crop.plot = TRUE, out.width = "0.5\\linewidth"}
plot(dune.bray.ord, display = "sites", type = "n")
```
\normalsize

## Adding other layers - axes scaling

\scriptsize
```{r NMDS20, fig.height = 5, crop.plot = TRUE, out.width = "0.5\\linewidth"}
plot(dune.bray.ord, display = "sites", type = "n")
points(dune.bray.ord, display = "sites", cex = 2)
```
\normalsize

## Adding other layers

\scriptsize
```{r NMDS21, fig.height = 5, crop.plot = TRUE, out.width = "0.5\\linewidth"}
plot(dune.bray.ord, display = "sites", type = "n")
points(dune.bray.ord,display = "sites", cex = 2)
ordispider(dune.bray.ord,groups = dune.env$Management, label = TRUE)
```
\normalsize

## Adding other layers

\scriptsize
```{r NMDS22, fig.height = 5, crop.plot = TRUE, out.width = "0.5\\linewidth"}
plot(dune.bray.ord, display = "sites", type = "n")
points(dune.bray.ord, display = "sites", cex = 2)
ordiellipse(dune.bray.ord,groups = dune.env$Management, label = TRUE)
```
\normalsize

## Adding other layers
\scriptsize
```{r NMDS23, fig.height = 5, crop.plot = TRUE, out.width = "0.5\\linewidth", message=FALSE, results='hide'}
plot(dune.bray.ord, display = "sites", type = "n")
points(dune.bray.ord,display = "sites", cex = 2)
ordisurf(dune.bray.ord,dune.env$A1, add = TRUE)
```
\normalsize

## Vectors in ordination space

\scriptsize
```{r NMDS24}
dune.bray.ord.A1.fit <- envfit(dune.bray.ord,dune.env$A1, permutations = 1000)
dune.bray.ord.A1.fit
```
\normalsize

## Vectors in ordination space

\scriptsize
```{r NMDS25, fig.height = 5, crop.plot = TRUE, out.width = "0.5\\linewidth", message=FALSE, results='hide'}
plot(dune.bray.ord, display = "sites", type = "n")
points(dune.bray.ord,display = "sites", cex = 2)
plot(dune.bray.ord.A1.fit, add = TRUE)
ordisurf(dune.bray.ord,dune.env$A1, add = TRUE)
```
\normalsize

